---
title: "P8105 Homework 3"
author: "Spencer Riddell"
output: github_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(janitor)
library(gt)
library(lubridate)
library(patchwork)
```

# Homework 3

---
  
  ## Problem 1
  
  ### 1. Instacart data description
  
```{r instacart-load}
library(p8105.datasets)
data("instacart")
glimpse(instacart)
summary(instacart)
head(instacart)
```

---
  
  ### 2. How many aisles are there, and which aisles are the most items ordered from?
  
```{r instacart-aisles}
aisle_counts <- instacart %>%
  count(aisle, sort = TRUE)

n_aisles <- n_distinct(instacart$aisle)
aisle_counts
n_aisles
```

---
  
  ### 3. Plot: Number of items ordered in each aisle (>10,000 items)
  
```{r instacart-aisle-plot, fig.height=8}
aisle_counts %>%
  filter(n > 10000) %>%
  mutate(aisle = fct_reorder(aisle, n)) %>%
  ggplot(aes(x = aisle, y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Items ordered by aisle (more than 10,000 items)",
    x = "Aisle",
    y = "Number of items ordered"
  )
```

---
  
  ### 4. Table: Three most popular items in selected aisles
  
```{r instacart-items-table}
selected_aisles <- c("baking ingredients", "dog food care", "packaged vegetables fruits")

instacart %>%
  filter(aisle %in% selected_aisles) %>%
  count(aisle, product_name, sort = TRUE) %>%
  group_by(aisle) %>%
  slice_max(order_by = n, n = 3) %>%
  ungroup() %>%
  arrange(aisle, desc(n)) %>%
  gt()
```

---
  
  ### 5. Table: Mean hour for Pink Lady Apples and Coffee Ice Cream by weekday
  
```{r instacart-mean-hour}
instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day), .groups = "drop") %>%
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour
  ) %>%
  gt()
```

---
  
## Problem 2
  
### 1. Import and clean Zillow data
  
```{r zillow-load}
zillow <- read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") %>% clean_names()
zipcodes <- read_csv("data/Zip Codes.csv") %>% clean_names()

# Standardize ZIP code column for joining
zillow <- zillow %>% rename(zip_code = region_name)
# Pivot to long format, parse date, join borough info
zillow_long <- zillow %>%
  pivot_longer(
    cols = matches("^\\d{4}_\\d{2}_\\d{2}$|^x\\d{4}_\\d{2}_\\d{2}$|^\\d{4}-\\d{2}-\\d{2}$"),
    names_to = "date",
    values_to = "rent"
  ) %>%
  mutate(
    date = str_replace_all(date, "^x", ""),
    date = str_replace_all(date, "_", "-"),
    date = as.Date(date)
  ) %>%
  left_join(zipcodes, by = "zip_code") %>%
  mutate(borough = county)

glimpse(zillow_long)
```

---
  
### 2. How many ZIP codes are observed 116 times? Fewer than 10 times? Why?
  
```{r zillow-summary}
zip_counts <- zillow_long %>%
  group_by(zip_code) %>%
  summarize(n_months = n())

n_zip_116 <- zip_counts %>% filter(n_months == 116) %>% nrow()
n_zip_lt10 <- zip_counts %>% filter(n_months < 10) %>% nrow()

n_zip_116
n_zip_lt10
```

*Add your commentary here.*
  
---
  
### 3. Table: Average rental price in each borough and year
  
```{r zillow-borough-year-table}
zillow_long <- zillow_long %>%
  mutate(year = year(date))

borough_year_table <- zillow_long %>%
  group_by(borough, year) %>%
  summarize(mean_rent = mean(rent, na.rm = TRUE)) %>%
  arrange(borough, year)

borough_year_table %>%
  pivot_wider(names_from = borough, values_from = mean_rent) %>%
  gt()
```

---
  
### 4. Plot: NYC Rental Prices within ZIP codes for all available years, by borough
  
```{r zillow-zip-plot, fig.height=6}
zillow_long %>%
  ggplot(aes(x = date, y = rent, group = zip_code, color = borough)) +
  geom_line(alpha = 0.2, size = 0.5) +
  facet_wrap(~borough) +
  labs(
    title = "NYC Rental Prices by ZIP Code (All Years)",
    x = "Date",
    y = "Rental Price"
  ) +
  theme_minimal()
```

---
  
### 5. Distribution of ZIP-code-level rental prices (2023) by borough
  
```{r zillow-2023-dist, fig.height=5}
zillow_2023 <- zillow_long %>%
  filter(year == 2023) %>%
  group_by(zip_code, borough) %>%
  summarize(mean_rent = mean(rent, na.rm = TRUE), .groups = "drop")

zillow_2023 %>%
  ggplot(aes(x = borough, y = mean_rent, fill = borough)) +
  geom_boxplot() +
  labs(
    title = "2023 ZIP Code Rental Price Distribution by Borough",
    x = "Borough",
    y = "Average ZIP Code Rental Price"
  ) +
  theme_minimal()
```

---
  
### 6. Combine previous two plots into a single graphic and export
  
```{r zillow-combined-plot, eval=FALSE}
p1 <- zillow_long %>%
  ggplot(aes(x = date, y = rent, group = zip_code, color = borough)) +
  geom_line(alpha = 0.2, size = 0.5) +
  facet_wrap(~borough) +
  labs(
    title = "NYC Rental Prices by ZIP Code (All Years)",
    x = "Date",
    y = "Rental Price"
  ) +
  theme_minimal()

p2 <- zillow_2023 %>%
  ggplot(aes(x = borough, y = mean_rent, fill = borough)) +
  geom_boxplot() +
  labs(
    title = "2023 ZIP Code Rental Price Distribution by Borough",
    x = "Borough",
    y = "Average ZIP Code Rental Price"
  ) +
  theme_minimal()

combined_plot <- p1 / p2
ggsave("results/zillow_combined.png", combined_plot, width = 12, height = 10)
```

---
  
## Problem 3
  
### 1. Load and tidy NHANES Accelerometer Data
  
```{r nhanes-load-tidy}
# Read data
demo <- read_csv("data/nhanes_covar.csv") %>% clean_names()
accel <- read_csv("data/nhanes_accel.csv") %>% clean_names()

# Recode sex and education as descriptive factors
demo <- demo %>%
  mutate(
    sex = case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female",
      TRUE ~ NA_character_
    ),
    education = case_when(
      education == 1 ~ "Less than high school",
      education == 2 ~ "High school equivalent",
      education == 3 ~ "More than high school",
      TRUE ~ NA_character_
    )
  )

# Filter: Age >= 21, no missing demo data
demo_clean <- demo %>%
  filter(age >= 21, !is.na(sex), !is.na(education))

# Merge datasets (joining by seqn)
nhanes <- left_join(demo_clean, accel, by = "seqn") %>%
  mutate(
    sex = factor(sex, levels = c("Male", "Female")),
    education = factor(
      education,
      levels = c("Less than high school", "High school equivalent", "More than high school")
    )
  )
```

---
  
### 2. Table: Number of Men and Women by Education
  
```{r nhanes-table-gender-education}
nhanes %>%
  count(sex, education) %>%
  pivot_wider(names_from = sex, values_from = n, values_fill = 0) %>%
  gt() %>%
  tab_header(title = "Number of Participants by Sex and Education")
```

---
  
### 3. Plot: Age Distributions for Men and Women by Education
  
```{r nhanes-age-dist, fig.height=4}
nhanes %>%
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) +
  facet_wrap(~education, ncol = 1) +
  labs(
    title = "Age Distribution by Sex and Education",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()
```

*Comment: The table and plot above show how participant demographics are distributed by sex and education.*
  
---
  
### 4. Total Daily Activity vs Age
  
```{r nhanes-total-activity}
nhanes <- nhanes %>%
  rowwise() %>%
  mutate(total_activity = sum(c_across(starts_with("min")), na.rm = TRUE)) %>%
  ungroup()

nhanes %>%
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_wrap(~education) +
  labs(
    title = "Total Daily Activity vs Age, by Sex and Education",
    x = "Age",
    y = "Total Daily Activity (MIMS sum)"
  ) +
  theme_minimal()
```

*Comment: The plot illustrates the relationship between age and total activity, stratified by sex and education.*
  
---
  
### 5. 24-Hour Activity Time Courses
  
```{r nhanes-activity-timecourse, fig.height=7}
nhanes_long <- nhanes %>%
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    names_prefix = "min",
    values_to = "activity"
  ) %>%
  mutate(
    minute = as.integer(minute),
    hour = minute %/% 60
  )

hourly_activity <- nhanes_long %>%
  group_by(education, sex, hour) %>%
  summarize(mean_activity = mean(activity, na.rm = TRUE), .groups = "drop")

hourly_activity %>%
  ggplot(aes(x = hour, y = mean_activity, color = sex)) +
  geom_line(size = 1) +
  facet_wrap(~education, ncol = 1) +
  labs(
    title = "Mean 24-Hour Activity by Education and Sex",
    x = "Hour of Day",
    y = "Mean MIMS"
  ) +
  theme_minimal()
```

*Comment: This plot shows the diurnal pattern of activity by education and sex. Look for differences in timing, amplitude, or overall activity levels.*
  
---